A portable, annotation-based, visual stepper for Common Lisp

### 2020-04-05 paper/paper.tex
Integrate review comments of third reviewer

A program stepper lets users execute code, section by section, in
steps of their own choosing.  A number of hidden control points are
inserted along the program's execution paths.  At each point, the
system may interrupt the program and wait for user instructions before
and/or after executing the next section.  Steppers often fall within
the category of \emph{program execution monitors}, along with
profilers, tracers and other debugging tools.  Stepping is the one of
the most popular forms of debugging since it allows users to study the
evolution of a program's state by direct inspection, as opposed to a
combination of conjecture and experimentation.  A stepper gives users
full control over the speed of the actual program and often even
allows manipulations of the program's state.  This frequently presents
a debugging advantage over guessing the program's state or monitoring
its outputs.

@@

Stepper systems usually function through text-based interfaces that
are capable of displaying the source code of the relevant code
sections.  Nevertheless, many users of steppers tool prefer to use
them through the more sophisticated user interfaces of editor programs
and integrated development environments (IDEs).  We shall call these
\emph{visual} steppers\footnote{After the nomenclature used in
  \cite{annotation-based}}: special-purpose programs running inside
the IDE communicate with the stepper tool by means of a protocol and
visually annotate the program's source text with the results of the
stepping session.  The expression about to be executed may e.g. be
marked with a red dot or highlighted in a special color.  Furthermore,
a modern user's expectations of a stepper system might include the
ability to add break points; to ``step over'' an expression; to ``step
into'' a call to a function defined elsewhere in the srouce code; to
``step out of'' the current call; to ``continue'' to a certain point,
and to inspect the values of local and global variables by
mouse-clicking on their manifestations in the source code.

@@

Common Lisp programmers are so detached from the practice of stepping
that some will simply declare that's proof that they don't need one at
all.  There is a hint of truth to this, as Common Lisp systems have
traditionally directed efforts to other types of debugging facilities
such as powerful interactive program restarts and function traces.  At
any rate, it seems undeniable that, stepper or no stepper, Common Lisp
programs will be debugged.

@@

Some Lisp implementations can and do implement stepping to various
levels of ability.  SBCL combines its implementation of
\texttt{CL:STEP} with the restarts system to provide a text-based
stepper, while the LispWorks\cite{lispworks-stepper} and Allegro CL
systems have sophisticated graphical stepping dialogues with views to
the source code and program state.  Unfortunately, these systems are
unavailable to whomever wishes to try portable programs across
different Common Lisp implementations.  Even if one switches between
two implementations that do have steppers, the difference in
capability and user interface is often enough to discourage either
system's use.

@@

Regrettably, even though Emacs provides visual stepper interfaces for
many programming languages, SLIME doesn't provide a portable stepper
interface.  We believe this to be due to the fact that the technical
challenges to be surmounted are greater than for other debugging
tools.  Among other problems, the portability mandate of SLIME implies
it is ultimately only allowed to invoke functionality present in all
the Common Lisp systems it connects to.  This means that even if all
implementations where to implement some form of \texttt{CL:STEP}, that
alone wouldn't be powerful enough to, say, communicate source location
information to and from Emacs, annotating the program source.

@@

In fact, the SLY Common Lisp IDE\cite{sly}, a derivative program of
SLIME, has redesigned some of its predecessor's underpinnings to make
this even easier.  We shall describe how ``stickers'', a feature that
SLY has recently acquired, lets users manually instrument selected
Common Lisp forms whose results they are interested in.  Stickers are
already a ``poor man's stepper'', in the sense that they have some
fundamental semantics of stepping but still encumber the user with
work than could be performed automatically.  To fill this gap, we
shall explore methods of combining stickers with automatic code
analysis.  We shall then be able to present an innovative stepper tool
for the Common Lisp language based on the SLY extension to the Emacs
editor, hereafter designated the \emph{SLY/Emacs stepper}.

@@

``Stepping is an old idea.''  So go the opening words of this
section's namesake in the article ``Annotation-Based Program
Stepping'', written by MIT's G. Parker in 1987\cite{annotation-based}.
In this article, the author surveys the 1970 decade's efforts to
develop various kinds of stepping tools in the MACLISP environment.
Likewise, we shall proceed to evaluate a small sample of Common Lisp
stepper systems, focusing on the ones that are portable, visual and
intersect our methodology.

@@

E.g. if the user marks the forms \texttt{(foo (bar))} and
\texttt{(bar)} inside the following expression:

@@

\begin{enumerate}
\item \texttt{Eclector}\cite{eclector}, a self-described ``portable
  Common Lisp reader that is highly customizable and can return
  concrete syntax trees'', is a full realization of a Common Lisp code
  reader that doesn't rely on the implementation's.  Its
  distinguishing characteristic are ``concrete syntax trees'' that
  aren't represented by \texttt{CONS} cells, rather by CLOS objects
  that mimic the properties of such cells while also keeping
  ``concrete'' source file information;

@@

\item A \emph{substitute read-table}, achieved by rebinding the
  variable \texttt{*READTABLE*}.  This table \emph{shadows} each of
  the entries of the current read-table (using
  \texttt{GET/SET-MACRO-CHARACTER}) with a function that allows full
  control of each characters influence in the returned symbolic
  expression.  By setting up this function in a particular manner, the
  resulting table can be made functionally indistinguishable from the
  original one, while retaining the ability to invoke a user's hook to
  record source positions.
\end{itemize}

@@

To reach a state where nothing but special and function forms exist, a
macroexpander must remove macros by expanding them.  However, in doing
so, our program \emph{must also remember} whence each macro's
expansion came, specifically the source position of the form in its
pre-expansion state.  This what we mean by \emph{mnesic
  macroexpansion}, the opposite of \emph{amnesic macroexpansion}

@@

These features don't seem hard to realize.  E.g. to enable more
sophisticated navigation behavior behavior the \texttt{RECORD} macro
could use a special variable to be made aware of recursive invocations
to itself or to the currently executing stack frame.  Thus we could
keep track of stickers being executed \emph{inside} each other,
i.e. within the dynamic scope of a previously active sticker
annotation.

@@

Common Lisp users are concerned about features that facilitate
day-to-day development.  \texttt{TRACE}, REPLs, \texttt{PRINT}
statements, the interactive debuggers, profilers and stickers are all
ways to solve certain debugging problems: some are more suitable to
some situations than others.  Therefore, it's important to note that
program stepping is just another tool in the toolbox, not a
panacea.

@@

By contrast, an hypothetical evaluation-based visual stepper (such an
as enhanced version of \cite{bourguignon}) would find it difficult to
maintain this advantage since the source-code correlation is achieved
at evaluation-time and is harder to mutate effectively afterwards.
Perhaps this fact can explain why the non-portable visual steppers of
the Allegro CL and LispWorks implementations don't work directly with
the source-editing facilities present in these
IDEs\footnote{LispWorks's stepper manual specifically mentions this
  fact in \cite{lispworks-stepper}.}.  A possible solution would be to
represent source-code correlation in terms of sub-expression paths in
the form tree instead of character positions, but it would still be
hard to resist deletions and insertions at top level.  A more
heavy-handed solution would lock the source file read-only for the
duration of the stepping session.  However, users are normally adverse
to such confinements.

>>>>

A program stepper lets users execute code, section by section, in
steps of their own choosing.  A number of hidden control points are
inserted along the execution paths of the program.  At each point, the
system may interrupt the program and wait for user instructions before
and/or after executing the next section.  Steppers often fall within
the category of \emph{program execution monitors}, along with
profilers, tracers and other debugging tools.  Stepping is the one of
the most popular forms of debugging since it allows users to study the
evolution of the state of a program by direct inspection, as opposed
to combining conjecture and experimentation.  A stepper gives users
full control over the speed of the actual program and often even
allows manipulations of that state.  This frequently presents a
debugging advantage over guessing a programs' state or monitoring its
outputs.

@@

Stepper systems usually function through text-based interfaces that
are capable of displaying the source code of the relevant code
sections.  Nevertheless, many users of stepper tools prefer to use
them through the more sophisticated user interfaces of editor programs
and integrated development environments (IDEs).  We shall call these
tools \emph{visual} steppers\footnote{After the nomenclature used in
  \cite{annotation-based}}: special-purpose programs running inside
the IDE communicate with the stepper tool by means of a protocol and
visually annotate the source text of a program with the results of the
stepping session.  The expression about to be executed may e.g. be
marked with a red dot or highlighted in a special color.  Furthermore,
a modern user's expectations of a stepper system might include the
ability to add break points; to ``step over'' an expression; to ``step
into'' a call to a function defined elsewhere in the source code; to
``step out of'' the current call; to ``continue'' to a certain point,
and to inspect the values of local and global variables by
mouse-clicking on their manifestations in the source code.

@@

Common Lisp programmers are so detached from the practice of stepping
that some will simply declare that's proof that they don't need one at
all.  There is a hint of truth to this declaration, as Common Lisp
systems have traditionally directed efforts to other types of
debugging facilities such as powerful interactive program restarts and
function traces.  At any rate, it seems undeniable that, stepper or no
stepper, Common Lisp programs will be debugged.

@@

Some Lisp implementations can and do implement stepping to various
levels of ability.  SBCL combines its implementation of
\texttt{CL:STEP} with the restarts system to provide a text-based
stepper, while the LispWorks\cite{lispworks-stepper} and Allegro CL
systems have sophisticated graphical stepping dialogues with views to
the source code and program state.  Unfortunately, these systems are
unavailable to whomever wishes to try portable programs across
different Common Lisp implementations.  Even if one switches between
two implementations that do have steppers, the difference in
capability and user interface is often enough to discourage the use of
either system.

@@

Regrettably, even though Emacs provides visual stepper interfaces for
many programming languages, SLIME doesn't provide a portable stepper
interface.  We believe this to be due to the fact that the technical
challenges to be surmounted are greater than for other debugging
tools.  Among other problems, the portability mandate of SLIME implies
it is ultimately only allowed to invoke functionality present in all
the Common Lisp systems it connects to.  This mandate implies that
even if all implementations where to implement some form of
\texttt{CL:STEP}, that alone wouldn't be powerful enough to, say,
communicate source location information to and from Emacs, annotating
the program source.

@@

In fact, the SLY Common Lisp IDE\cite{sly}, a derivative program of
SLIME, has redesigned some of the underpinnings of its predecessor to
make the development of new extensions easier.  We shall describe how
``stickers'', a feature that SLY has recently acquired, lets users
manually instrument selected Common Lisp forms whose results they are
interested in.  Stickers are already a ``poor man's stepper'', in the
sense that they have some fundamental semantics of stepping but still
encumber the user with work than could be performed automatically.  To
fill this gap, we shall explore methods of combining stickers with
automatic code analysis.  We shall then be able to present an
innovative stepper tool for the Common Lisp language based on the SLY
extension to the Emacs editor, hereafter designated the
\emph{SLY/Emacs stepper}.

@@

``Stepping is an old idea.''  So go the opening words of this
section's namesake in the article ``Annotation-Based Program
Stepping'', written by MIT's G. Parker in 1987\cite{annotation-based}.
In this article, the author surveys the efforts of the 1970 decade to
develop various kinds of stepping tools in the MACLISP environment.
Likewise, we shall proceed to evaluate a small sample of Common Lisp
stepper systems, focusing on the ones that are portable, visual and
intersect our methodology.

@@

E.g., if the user marks the forms \texttt{(foo (bar))} and
\texttt{(bar)} inside the following expression:

@@

\begin{enumerate}
\item \texttt{Eclector}\cite{eclector}, a self-described ``portable
  Common Lisp reader that is highly customizable and can return
  concrete syntax trees'', is a full realization of a Common Lisp code
  reader that doesn't rely on any preexisting reader.  Its
  distinguishing characteristic are ``concrete syntax trees'' that
  aren't represented by \texttt{CONS} cells, rather by CLOS objects
  that mimic the properties of such cells while also keeping
  ``concrete'' source file information;

@@

\item A \emph{substitute read-table}, achieved by rebinding the
  variable \texttt{*READTABLE*}.  This table \emph{shadows} each of
  the entries of the current read-table (using
  \texttt{GET/SET-MACRO-CHARACTER}) with a function that fully
  controls the influence of each character over the returned symbolic
  expression.  By setting up this function in a particular manner, the
  resulting table remains functionally indistinguishable from the
  original one, while gaining the ability to invoke a hook that
  records source positions.
\end{itemize}

@@

To reach a state where nothing but special and function forms exist, a
macroexpander must remove macro calls by expanding them.  However, in
doing so, our program \emph{must also remember} whence each macro's
expansion came, specifically the source position of the form in its
pre-expansion state.  This behavior is what we refer to as
\emph{mnesic macroexpansion}, the opposite of \emph{amnesic
  macroexpansion}.

@@

These features don't seem hard to realize.  E.g., to enable more
sophisticated navigation behavior behavior the \texttt{RECORD} macro
could use a special variable to be made aware of recursive invocations
to itself or to the currently executing stack frame.  Thus we could
keep track of stickers being executed \emph{inside} each other,
i.e. within the dynamic scope of a previously active sticker
annotation.

@@

Common Lisp users are concerned about features that facilitate
day-to-day development.  \texttt{TRACE}, REPLs, \texttt{PRINT} forms,
the interactive debuggers, profilers and stickers are all ways to
solve certain debugging problems: some are more suitable to some
situations than others.  Therefore, it's important to note that
program stepping is just another tool in the toolbox, not a panacea.

@@

By contrast, a hypothetical evaluation-based visual stepper (such an
as enhanced version of \cite{bourguignon}) would find it difficult to
maintain this advantage since the source-code correlation is achieved
at evaluation-time and is harder to mutate effectively afterwards.
Perhaps this fact can explain why the non-portable visual steppers of
the Allegro CL and LispWorks implementations don't work directly with
the source-editing facilities present in these
IDEs\footnote{LispWorks's stepper manual specifically mentions this
  fact in \cite{lispworks-stepper}.}.  A possible solution would be to
represent source-code correlation in terms of sub-expression paths in
the form tree instead of character positions, but it would still be
hard to resist deletions and insertions at top level.  A more
heavy-handed solution would lock the source file read-only for the
duration of the stepping session.  However, users are normally adverse
to such confinements.
